NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2020-12-10 Manuscript/manuscriptPGPE.tex
Update intro

\begin{abstract}
% insert abstract here
Abstract
\end{abstract}

@@

Lipid composition of biomembranes vary greatly between different
organelles and organism, and different lipid types interact with
proteins in many different ways regulating their function \cite{??}.
Lipids may directly interact with proteins by specific binding sites
or indirectly affect protein functions by altering membrane properties
or charge \cite{??}. Specific interactions with certain lipid headgroups
are known to be essential for the function of several proteins \cite{??}.
However, it is not clear if the specific binding is driven by
the differences in acessible conformational states between lipid types or
by specific intermolecular interactions between lipids and proteins.

Differences between preferred conformations of different lipid types
are observed in lipid crystal structures \cite{buldt81,pascher92},
but lipids are not in solid state in biomembranes. 
On the other hand, NMR experiments enable detection of lipid headgroups
in physiological liquid states also in cells \cite{gally81,scherer87,seelig90}.
These studies suggest that the glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, 
glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90},
and the headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. 
%Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
%In contrast to PC lipids (see \cite{botan15} and references therein).
%Some attempts to resolve conformational ensembles from NMR for PC and PE lipids have been made,
%but lesser extend for PG or PS lipids \cite{seelig77c,davis83,Semchyschyn04}.
Attempts to resolve the conformational ensembles of lipid headgroup rely mainly on
absolute values of C-H bond order parameters measured using $^2$H NMR~\cite{seelig77c,davis83,Semchyschyn04}.
However, appropriate models to interpret the experimental data have not been available and
also the signs of the order parameters are required to fully capture the details in conformational ensembles  \cite{botan15,ollila16}.
%determine the structural 
%Classical molecular dynamics simulations could potentially give such ensembles and therefore enable
%the detailed studies of lipid headgroup behaviour in complex biomolecular systems, but current
%force fields are not accurate enough to reproduce the correct conformational ensembles for PC and PS headgroups \cite{botan15,antila19}.

PE and PG lipids are most common lipids in bacteria \cite{sohlenkamp16}.
Zwitterionic PE is the second most abundant glycerophospholipid in eukaryotic cells
and has been related to the diseases \cite{vance15,calzada16,patel17}.
Anionic PG lipids are less abundant, but is also proposed to be fundamental for terrestrial life \cite{furse17}.
PE and PG affect membrane protein functionality \cite{hariharan18} and bind to various proteins \cite{yeagle14}.
PE headgroup is also prone for negative membrane curvature and causes membrane fusion \cite{Chernomordik08,calzada16}.
Therefore, the PE and PG headgroup structures play probably essential roles in 
many biological processes.

@@

Derivation of different lipid headgroup conformational ensembles in liquid state has been
inconclusive due to lack of suitable experimental data and tools to interpret the conformational ensembles.

Besides the structure, also ion binding may regulate biophysical activity of especially negatively
charged lipid headgroups \cite{seelig90}. Monovalent cation (except Lithium) binding to zwitterionic PC and anionic PS headgroups is
very weak, while multivalent ion binding is stronger but still weak \cite{cevc90,tocanne90,roux90,catte16,antila19}. The ion binding affinity
data for PE is more scarce \cite{marra85}, but large differences to PC would be surprising.
Negatively charged lipids are suggested to bear same cation binding constants than zwitterionic lipids, but
the amount of bound ions to negatively charged membranes would still be larger because
the concentration of cations in the vicinity of membranes would be higher  \cite{seelig90}.
%is stronger
%Based on the electrometer concept and other data is has been suggested that \cite{seelig90}
%\begin{displayquote}
%  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
%    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
%    (ii) the free Ca$^{2+}$
%    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
%    charge, either due to protein or to lipid; \\
%    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
%    of bound Ca$^{2+}$ at neutral and charged lipids; \\
%    (iv) the actual binding step can be described by a Langmuir
%    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
%    used to describe the chemical binding equilibrium.''}
%\end{displayquote}
On the other hand, anionic PS lipids are proposed chelate with calcium ions \cite{hauser85,feigenson86,roux91}.
In simulations, the cation binding affinity to PC and PS membranes is typically overestimated \cite{catte16,antila19},
which can be improved by applying the ECC to the partial charges of the force fields \cite{melcr18,melcr19}.

Here, we use open collaboration and order parameters of glycerol backbone and headgroup
to evaluate the accuracy of PE and PG heagroup structures, and the cation binding affinity

to anionic membranes containing PG lipids in the current MD simulation force fields.
The force field giving the best description for glycerol backbone and headgroup structures
of PC, PS, PG and PE headgroups (CHARMM36) reproduces the essential differences in order parameters
between these headgroups, and therefore enables the analysis of structural differences between the
headgroups.

>>>>

\begin{abstract}
% insert abstract here
The force field giving the best description for glycerol backbone and headgroup structures
of PC, PS, PG and PE headgroups (CHARMM36) reproduces the essential differences in order parameters
between these headgroups, and therefore enables the analysis of structural differences between the
headgroups.

\end{abstract}

@@

Chemical compositions of hydrophilic lipid headgroups vary between different
organelles and organisms, and different lipid types
regulate protein functions in many different ways \cite{lee03,vanmeer08}.
Lipids can directly bind to proteins or indirectly affect protein
functions by altering membrane properties or charge \cite{lee03,lemmon08}.
While the specific interactions with certain lipid headgroups
are known to be essential for the function of several proteins \cite{lee11,lemmon08},
it is not clear if the specificity is driven by the differences in accessible
conformational states between lipid types or
by specific intermolecular lipid-protein interactions.

Lipid conformational ensembles in physiologically relevant lamellar liquid phase
are typically derived from NMR experiments, particularly from 
%Attempts to resolve the conformational ensembles of lipid headgroup rely mainly on
C-H bond order parameters measured using $^2$H NMR~\cite{seelig77c,davis83,Semchyschyn04}
which can be detected even from cells \cite{gally81,scherer87,seelig90}.
%but lipids are not in solid state in biomembranes. 
%On the other hand, NMR experiments enable detection of lipid headgroups
%in physiological liquid states also in cells \cite{gally81,scherer87,seelig90}.
These studies suggest that the glycerol backbone structures are largely similar irrespectively of the headgroup \cite{gally81}, 
and the headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. 
%Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
%In contrast to PC lipids (see \cite{botan15} and references therein).
%Some attempts to resolve conformational ensembles from NMR for PC and PE lipids have been made,
%but lesser extend for PG or PS lipids \cite{seelig77c,davis83,Semchyschyn04}.
However, these results are based only on the absolute values while 
%we have recently demonstrated that also
the necessity of order parameters signs in capturing the lipid conformational ensembles has been resently demonstrated \cite{botan15,ollila16,ferreira16}.
Furthermore, the detailed understanding of lipid conformational ensembles
is limited by the lack of universal models to interpret the experiments \cite{pezeshkian18,akutsu20}.
%do not exist and
%the proposed structures are often based on lipid crystal structures \cite{pezeshkian18,akutsu20}.
%which were solved many decades ago \cite{buldt81,pascher92}.
%determine the structural 
%Classical molecular dynamics simulations could potentially give such ensembles and therefore enable
%the detailed studies of lipid headgroup behaviour in complex biomolecular systems, but current
%force fields are not accurate enough to reproduce the correct conformational ensembles for PC and PS headgroups \cite{botan15,antila19}.

Stuctures of different lipid types in protein bound states can be extracted from the protein data bank (PDB) \cite{berman00},
but their relation to the conformational ensembles in liquid lamellar state remains unclear \cite{marsh13b}.
In addition to the changes in lipid conformational ensembles upon binding to proteins,
%PC headgroup conformations do not change upon addition of zwitterionic lipids or cholesterol \cite{scherer87,ferreira13} into a membrane,
%but are sensitive to the presence of charged molecules in membranes \cite{seelig87}.
also the experimentally measured response of lipid headgroup to membrane bound charges remains poorly understood 
%The response to charges is explained by the change in the orientation of lipid headgroup dipole vector due to electrostatic interactions,
%but the detailed changes are not yet understood
due to the lack of suitable models to interpret the lipid conformational ensembles in liquid lamellar state \cite{Semchyschyn04}.

Here, we use natural abundance $^{13}$C NMR experiments and MD simulations from the NMRlipids open collaboration
to resolve the differences in conformational ensembles between PC, PE, PG and PS lipid headgroups.
%open collaboration and order parameters of glycerol backbone and headgroup
%to evaluate the accuracy of PE and PG heagroup structures, and the cation binding affinity
Zwitterionic PC and PE are the most common lipids in eukaryotes and bacteria, respectively \cite{vanmeer08,sohlenkamp16}.
PE is also the second most abundant glycerophospholipid in eukaryotic cells
and has been related to various diseases \cite{vance15,calzada16,patel17}.
PS and PG are the most common negatively charged lipids in eukaryotes and bacteria, respectively,
and affect membrane protein functionality and signaling \cite{lemmon08,leventis10,sohlenkamp16,hariharan18}.
%PE and PG lipids are most common lipids in bacteria \cite{sohlenkamp16}.
%Anionic PG lipids are less abundant, but is also proposed to be fundamental for terrestrial life \cite{furse17}.
%PE and PG affect membrane protein functionality \cite{hariharan18} and
All the studied lipids specifically bind to various proteins \cite{yeagle14}.
%PE headgroup is also prone for negative membrane curvature and causes membrane fusion \cite{Chernomordik08,calzada16}.
%Therefore, the PE and PG headgroup structures play probably essential roles in 
%many biological processes.
Furthermore, we use our results to elucidate lipid-protein interactions and the effect of charges on lipid conformations.

The lipid conformational ensembles in liquid lamellar state paves the way toward understanding the
specific binding of different lipid types to membrane proteins and how they regulate the protein function.
Because glycerol backbone and headgroup structures of PC lipids are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90},
the results from model systems could be used to understand the biological role of lipid headgroup conformational ensembles in different lipid types.

@@

%Derivation of different lipid headgroup conformational ensembles in liquid state has been
%inconclusive due to lack of suitable experimental data and tools to interpret the conformational ensembles.

%Besides the structure, also ion binding may regulate biophysical activity of especially negatively
%charged lipid headgroups \cite{seelig90}. Monovalent cation (except Lithium) binding to zwitterionic PC and anionic PS headgroups is
%very weak, while multivalent ion binding is stronger but still weak \cite{cevc90,tocanne90,roux90,catte16,antila19}. The ion binding affinity
%data for PE is more scarce \cite{marra85}, but large differences to PC would be surprising.
%Negatively charged lipids are suggested to bear same cation binding constants than zwitterionic lipids, but
%the amount of bound ions to negatively charged membranes would still be larger because
%the concentration of cations in the vicinity of membranes would be higher  \cite{seelig90}.
%
%is stronger
%Based on the electrometer concept and other data is has been suggested that \cite{seelig90}
%\begin{displayquote}
%  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
%    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
%    (ii) the free Ca$^{2+}$
%    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
%    charge, either due to protein or to lipid; \\
%    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
%    of bound Ca$^{2+}$ at neutral and charged lipids; \\
%    (iv) the actual binding step can be described by a Langmuir
%    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
%    used to describe the chemical binding equilibrium.''}
%\end{displayquote}
%
%On the other hand, anionic PS lipids are proposed chelate with calcium ions \cite{hauser85,feigenson86,roux91}.
%In simulations, the cation binding affinity to PC and PS membranes is typically overestimated \cite{catte16,antila19},
%which can be improved by applying the ECC to the partial charges of the force fields \cite{melcr18,melcr19}.
